R Notebook

### 2017-03-02 manuscript/draft.Rmd
Edits to Intro/Box in response to @davharris's suggestions

# Moving towards large-scale data-intensive forecasting of species richness in breeding birds

Authors: Harris, Taylor and White?

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and testing the performance of ecological models
[@clark2001, @dietz2017]. One area of particular interest for forecasting is
biodiversity since it is important for ecosystem function, central to
conservation planning, and expected to be influenced by anthropogenic global
change. Anticipating potential changes in biodiversity is crucial for large
scale management and conservation, and for addressing debates regarding whether
local scale diversity is declining and may continue to decline in response to
anthropogenic influences [@dornelas2014].

@@

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. What we do know comes from a small number of hindcasting
studies, where models are built using data on species occurrence and richness
from the past and evaluated on their ability to predict contemporary patterns
[e.g., @algar2009, @distler2015]. These studies are a valuable first step, but
lack several components that are important for developing good forecasting
models and understanding how accurately these models can predict the
future. These components of good forecasting and evaluation (Box 1) broadly
involve: 1) expanding the use of data to include biological and environmental
time-series [@treddenick2016]; 2) accounting for uncertainty in observations,
processes, model choice, and forecast evaluation [@dietz2017]; and 3) conducting
meaningful evaluations of the forecasts by hindcasting, archiving short-term
forecasts, and comparing forecasts to baselines to determine if the forecasts
are more accurate than assuming the system is basically static [@Perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. We compare traditional distribution modeling
based approaches to spatial models of species richness, time-series methods, and
simple baselines. All forecasts model uncertainty and observation, are evaluated
across different time lags using hindcasting, and are publicly archived to allow
future assessment. We discuss the implications of these practices for our
understanding of, and confidence in, the resulting forecasts, and how we can
continue to improve on these approaches in the future.

@@

Typically ecological forecasts use one modeling approach or a small number of
related approaches. By fitting and evaluating multiple modeling approaches we
can learn more rapidly about the best approaches for making predictions for a
given ecological quantity. This includes comparing process based and data-driven
models.

@@

Understanding confidence in a forecast is just as important as understanding the
average or expected outcome. Failing to account for uncertainty can result in
overconfidence in highly uncertain outcomes leading to poor decision making and
erosion of confidence in ecological forecasts. Models should explicitly include
sources of uncertainty where possible and evaluations of forecasts should assess
the accuracy of uncertainties as well as point estimates [@dietz2017].

@@

In cases where long-term predictions are the primary goal, short-term should
also be made to accommodate the time-scales of planning and management decisions
and to allow the accuracy of the forecasts to be quickly evaluated
[@treddenick2016a];

### 10. Compare forecasts to simple baselines

Compare the accuracy of the forecasts to simple baselines such as the
time-series average to determine if the modeled forecasts are more accurate than
the naive assumption that the world is static.

>>>>

---
output:
  pdf_document: default
  html_document: default
---
# Moving towards large-scale data-intensive forecasting of species richness in breeding birds
# Using best practices for forecasting biodiversity changes in breeding birds
Authors: Harris, Taylor and White?

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001, @dietz2017]. One area of
particular interest for forecasting is biodiversity since it is important for
ecosystem function, central to conservation planning, and expected to be
influenced by anthropogenic global change. Anticipating potential changes in
biodiversity is crucial for large scale management and conservation, and for
addressing debates regarding whether local scale diversity is declining and may
continue to decline in response to anthropogenic influences [@dornelas2014].

@@

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009, @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@treddenick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietz2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@Perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. This 30 year time-scale has only recently
become possible for large-scale time-series based forecasting and allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and simple baselines. All forecasts model
uncertainty and observation, are evaluated across different time lags using
hindcasting, and are publicly archived to allow future assessment. We discuss
the implications of these practices for our understanding of, and confidence in,
the resulting forecasts, and how we can continue to improve on these approaches
in the future.

@@

Typically ecological forecasts use one modeling approach or a small number of
related approaches. By fitting and evaluating multiple modeling approaches we
can learn more rapidly about the best approaches for making predictions for a
given ecological quantity. This includes comparing process based and data-driven
models and comparing the accuracy of forecasts to simple baselines to determine
if the modeled forecasts are more accurate than the naive assumption that the
world is static.

@@

Understanding confidence in a forecast is just as important as understanding the
average or expected outcome. Failing to account for uncertainty can result in
overconfidence in highly uncertain outcomes leading to poor decision making and
erosion of confidence in ecological forecasts. Models should explicitly include
sources of uncertainty and how the propagate through the forecast where
possible. Evaluations of forecasts should assess the accuracy of uncertainties
as well as point estimates [@dietz2017].

@@

In cases where long-term predictions are the primary goal, short-term should
also be made to accommodate the time-scales of planning and management decisions
and to allow the accuracy of the forecasts to be quickly evaluated
[@treddenick2016a].
